# Hockey

## R Packages

```{r}
library(xml2)
library(rvest)
library(janitor)
library(dplyr)
library(sportyR)
library(ISAR)
library(ggplot2)
```

```{r}
library(scales) # used for scaling data and providing good breaks and labels for graphs
library(stringr) # helps when cleaning and working with string data
library(hockeyR) # play-by-play data for NHL games
```

## If you don’t know anything about hockey 

## What you will need to complete this chapter

```{r}
#| eval: false

year <- "2023"
link <- paste("http://moneypuck.com/moneypuck/playerData/seasonSummary/",
              year, 
              "/regular/skaters.csv",
              sep = "")
df <- read.csv(link)
```

## Exploring Ice Hockey through Basic and Advanced Statistics

```{r}
years <- 2012:2023
for(i in seq_along(years)){
  cat(sprintf('Year: %s at %s\n', years[i], Sys.time()))
  reg_url <- paste("https://www.hockey-reference.com/leagues/NHL_", years[i],
                   "_skaters.html", sep = "")
  pg <- read_html(reg_url)
  nhl_reg_stats <- rvest::html_table(pg, fill = TRUE)[[1]]
  names(nhl_reg_stats) <- nhl_reg_stats[1, ]
  nhl_reg_stats_2 <- janitor::clean_names(nhl_reg_stats) |> 
    filter(rk != "Rk")
  
  adv_url <- paste("https://www.hockey-reference.com/leagues/NHL_", years[i],
                   "_skaters-advanced.html", sep = "")
  pg_2 <- read_html(adv_url) 
  nhl_adv_stats <- rvest::html_table(pg_2, fill = TRUE)[[1]]
  names(nhl_adv_stats) <- nhl_adv_stats[1, ]
  nhl_adv_stats_2 <- janitor::clean_names(nhl_adv_stats) |> 
    filter(rk != "Rk")
    
  tmp_results <- dplyr::inner_join(nhl_reg_stats_2, nhl_adv_stats_2) |> 
    mutate(season = years[i]) |> 
    select(-rk)
  
  if(exists("nhl_data")){
    nhl_data <- dplyr::bind_rows(nhl_data, tmp_results)
  }
  else(nhl_data <- tmp_results)
  
}
```

```{r}
nhl_data[, -c(1, 3, 4, 22, 41, 42)] <- sapply(nhl_data[, -c(1, 3, 4, 22, 41, 42)], 
                                             as.numeric )
```

## *Corsi* and *Fenwick* metrics

```{r}
mac <- filter(nhl_data, player == "Nathan MacKinnon")
```

```{r}
#| label: fig-mac-corsi
#| fig-cap: "Corsi percentage values (y) by age (x) for Nathan MacKinnon of the Colorado Avalanche."

p <- ggplot(data = mac,
            aes(x = age, y = cf_percent/100))
p + geom_line() +
  geom_point() +
  labs(x = "Age", y = "Corsi Percentage") +
  scale_y_continuous(labels = scales::percent) +
  scale_x_continuous(breaks = 18:28) +
  theme_bw()
```

## Time On Ice (TOI) by Positions

```{r}
#| label: fig-toi
#| fig-cap: "Density plots for Time on Ice by position."

p <- ggplot(data = nhl_data |> 
              mutate(toi = toi) |> 
              filter(toi >= 200, 
                     !(pos %in% c("F", "W"))),
            aes(x = toi))
p + geom_density(alpha = .25, fill = "grey20") +
  facet_wrap(~ pos, nc = 2) +
  labs(x = "Time on Ice (min)") +
  theme_bw()
```

## Evaluating Players

```{r}
top_cor <- nhl_data |> 
  filter(toi >= 200) |> 
  group_by(season, pos) |> 
  summarize(max_corsi = max(cf_percent)) |> 
  arrange(desc(max_corsi))
head(top_cor)
```

```{r}
top_cor <- nhl_data |>
  filter(toi >= 200) |>
  group_by(season, pos) |>
  filter(cf_percent == max(cf_percent)) |>
  select(season, player, pos, tm, cf_percent) |>
  arrange(desc(cf_percent))
head(top_cor)
```

```{r}
#| label: fig-corsi-pos
#| fig-cap: "Average Corsi values over time by several positions."

nhl_tmp <- nhl_data |>
  filter(toi >= 200) |>
  group_by(season, pos) |>
  summarize(avg_corsi = mean(cf_percent))
p <- ggplot(nhl_tmp,
            aes(x = season, y = avg_corsi, col = pos))
p + geom_line() +
  geom_point() +
  labs(y = "Average Corsi",
       x = "Season",
       linetype = "Position") +
  scale_x_continuous(breaks = 2008:2023) +
  scale_color_brewer("Position", palette = "Dark2") +
  theme_bw()
```

## Research Block: Fighting in the NHL
## Accessing the NHL API

```{r}
pbp_23 <- load_pbp('2022-23')
```

```{r}
avs_pbp <- pbp_23 |> 
  filter(event_team_abbr == "COL")
```

```{r}
unique(avs_pbp$event)
```

```{r}
avs_goals <- avs_pbp |> 
  filter(event == "Goal")
```

```{r}
#| label: fig-nhl-avs-game-goals
#| fig-cap: "All shot locations with goals in Colorado Avalanche games from the 2022-23 season."

nhl_rink <- geom_hockey('nhl')
nhl_rink +
  geom_point(data = avs_goals, 
             aes(x = x_fixed, y = y_fixed), 
             alpha = .5) +
  labs(caption = "")
```

```{r}
avs_goals_playoffs <- avs_goals |> 
  filter(substr(game_id, 5, 6) == "03")
```

```{r}
#| label: fig-nhl-avs-playoff-goals
#| fig-cap: "All shot locations with goals in Colorado Avalanche playoff games from the 2022-23 season."

nhl_rink <- geom_hockey('nhl')
nhl_rink +
  geom_point(data = avs_goals_playoffs, 
             aes(x = x_fixed, y = y_fixed), 
             alpha = 1) +
  labs(caption = "")
```



## Shiny Application {#sec-shiny}

```{r}
#| eval: false
#| include: true

library(shiny)
library(ISAR)
library(ggplot2)
library(plotly)
```

```{r}
#| eval: false
#| include: true

df <- nhl_team_stats_2022 |> 
  select(starts_with("stat"), team_name) |> 
  rename_with(~ sub("stat_", "", .x), everything())
```


```{r}
#| eval: false
#| include: true

ui <- fluidPage(

    # Application title
    titlePanel(div(HTML("Shiny Example for <em>Introduction to 
                        Sports Analytics using R</em>"))),
    # Sidebar with a slider input for number of bins 
    sidebarLayout(
        sidebarPanel(
          selectInput("selector_x", h4("X Variable"), 
                      choices = names(df)[2:31], 
                      selected = "wins", 
                      width = '190px'),
          selectInput("selector_y", h4("Y Variable"), 
                      choices = names(df)[2:31], 
                      selected = "losses", 
                      width = '190px'),
          width = 3
          ),

        # Show a plot of the generated distribution
        mainPanel(
          plotlyOutput("graph")
        )
    )
)
```

```{r}
#| eval: false
#| include: true

server <- function(input, output, session) {

  output$graph <- renderPlotly({
    ggplotly(ggplot(data = df,
                    aes_string(x = input$selector_x,
                               y = input$selector_y),
                    text = paste('X: ', input$selector_x, '\n',
                                 'Y: ', input$selector_y, sep = "")) +
               geom_point(aes(col = team_name)) +
               theme_bw() +
               labs(x = input$selector_x,
                    y = input$selector_y,
                    colour = "Teams"),
             tooltip = "text")
  })
}

# Run the application 
shinyApp(ui = ui, server = server)
```

## Case Study: Pulling the goaltender


## Exercises

