# Sport Wagering and Daily Fantasy Sports

## R Packages

```{r}
library(ISAR)
library(remotes)
library(dplyr)
library(janitor)
library(bettoR)
library(lpSolve)
library(scales)
library(ggplot2)
```

```{r}
library(bettoR) #functions for wagering operations and calculations
library(lpSolve) # solves linear, integer and mixed integer programs
```

```{r}
#| eval: false

remotes::install_github("a-i-sports/bettoR")
```

## If you don’t know anything about Sports Wagering


## Sports Wagering and Morality

## Sports Wagering and Legality

## Basics of Sports Wagering

## The role of the bookmaker

```{r}
#Federer Wins or Nadal Wins
federer_exposure = 300
nadal_exposure = 123.08
bookmaker_profit_federer_wins = nadal_exposure - 
  (federer_exposure/3) #federer backers had to lay 3 to 1 odds
bookmaker_profit_nadal_wins = federer_exposure - 
  (nadal_exposure * 2.25) # nadal backers receive 2.25 to 1 odds
cat("When Federer wins, the bookmaker profits", 
    bookmaker_profit_federer_wins)
cat("When Nadal wins, the bookmaker profits", 
    bookmaker_profit_nadal_wins)
```

```{r}
# Nadal Wins
300-(50*2.25)
# Federer Wins
50-(300/3)
```

## Popular Types of Wagers

## Using Analytics to Set Market Prices
## Displaying and Calculating Market Prices

```{r}
#| label: tbl-equiv-odds
#| tbl-cap: Hypothetical tennis match odds in the various formats discussed in this chapter.  
#| echo: false

df <- data.frame(`Winner` = c("Federer", "Nadal"),
                 `American` = c("-300", "+225"),
                 `Decimal` = c(1.33, 3.25),
                 `Fractional` = c(1/3, 9/4),
                 `Probability` = c(0.75, 0.30))
df |> 
  gt() |> 
  opt_row_striping()
```


```{r}
implied_odds(0.75, type = "all")
```

```{r}
hold_calc(-110, -110)
```

## Straight Wagers vs. Parlays
```{r}
side_wager = 10
side_multiplier = 1.53 # -190 in decimal odds
total_wager = 10
total_multiplier = 1.91 #-110 in decimal odds
payout = side_wager * side_multiplier + total_wager * total_multiplier
profit = payout - side_wager - total_wager
profit
```

```{r}
convert_odds(-190)
convert_odds(-110)
```

```{r}
convert_odds(-190)[1]*convert_odds(-110)[1]
```

```{r}
convert_odds(-190)[1]*convert_odds(-110)[1]*20
```

```{r}
convert_odds(-190)[1]*convert_odds(-110)[1]*20 - 20
```

## Research Block 

## Kelly Criterion


```{r}
0.6 - (0.4/1)
```

```{r}
0.55 - (0.45/0.91)
```

```{r}
kelly(win_prob = 0.55, odds = -110, type = "us")
kelly(win_prob = 0.55, odds = 1.91, type = "dec")
```

```{r}
kelly(win_prob = 0.54, odds = -110, type = "us")
```

```{r}
kelly(win_prob = 0.53, odds = -110, type = "us")
kelly(win_prob = 0.52, odds = -110, type = "us")
```

## Case Study: Stevin Smith, Tim Donaghy, and the issues of sports fixing 

## Daily Fantasy Sports

```{r}
#| label: tbl-dk-skaters-pts
#| tbl-cap: Draftkings points (skaters) for an NHL showdown game. 
#| echo: false

df <- data.frame(Statistics = c("Goal", "Assist", "Shot on Goal", "Blocked Shot",
                                "Shorthanded Goal", "Faceoff Won", "Shootout Goal",
                                "Hat Trick"),
                 Points = c(8, 5, 2, 1.1, 2, 0.2, 1.5, 3))
df |> 
  gt::gt() |> 
  gt::opt_row_striping()
```

```{r}
#| label: tbl-dk-goalies-pts
#| tbl-cap: Draftkings points (goalies) for an NHL showdown game. 
#| echo: false

df <- data.frame(Statistics = c("Win", "Save", "Goal Allowed", "Shutout"),
                 Points = c(4, 0.5, -1, 4))
df |> 
  gt::gt() |> 
  gt::opt_row_striping()
```

```{r}
df <- dk_nyr_car |> 
  filter(roster_position == "FLEX") #be sure we can include any player
```

```{r}
obj <- df$avg_points_per_game
```

```{r}
cons <- rbind(rep(1, nrow(df)),                       # player constraint
              df$salary,                              # salary constraint
              t(model.matrix(~ team_abbrev + 0, df))) # team constraints
cons[1:4, 1:8] # to inspect the matrix
```

```{r}
dir <- c("=", "<=", ">=", ">=")
```

```{r}
rhs <- c(6, 50000, 1, 1)
```

```{r}
result <- lpSolve::lp("max", obj, cons, dir, rhs, all.bin = TRUE)
result
```

```{r}
team <- df[which(result$solution == 1), ] |> 
  select(name, position, salary, team_abbrev)
team
```

```{r}
teams_mat <- t(model.matrix(~ team_abbrev + 0, df))
goalies <- ifelse(df$position == "G", 1, 0)
teams_goalies <- t(goalies * t(teams_mat))

cons_tmp <- rbind(cons, teams_goalies)
dir_tmp <- c(dir, "<=", "<=")
rhs_tmp <- c(rhs, 1, 1)
result <- lp("max", obj, cons_tmp, dir_tmp, rhs_tmp, all.bin = TRUE)
team <- df[which(result$solution == 1), ] |> 
  select(name, position, salary, team_abbrev, avg_points_per_game)
team
```

```{r}
sum(team$salary)
```

```{r}
df_new <- df |>  
  filter(position != "G" | name %in% c("Antti Raanta", "Igor Shesterkin"))
obj <- df_new$avg_points_per_game
cons <- rbind(rep(1, nrow(df_new)), # player constraints
              df_new$salary, # salary constraints
              t(model.matrix(~ team_abbrev + 0, df_new))) # team constraints
dir <- c("=", "<=", ">=", ">=")
rhs <- c(6, 50000, 1, 1)
result <- lp("max", obj, cons, dir, rhs, all.bin = TRUE)
team <- df_new[which(result$solution == 1), ] |> 
    select(name, position, salary, team_abbrev, avg_points_per_game)
team
```

```{r}
results <- data.frame(P1 = rep(NA, 100),
                      P2 = rep(NA, 100),
                      P3 = rep(NA, 100),
                      P4 = rep(NA, 100),
                      P5 = rep(NA, 100),
                      P6 = rep(NA, 100),
                      points = rep(NA, 100),
                      salary = rep(NA, 100))
results[1, ] <- c(team$name, sum(team$avg_points_per_game), sum(team$salary))
for (i in 1:99){ ## 99 will give us 100 total teams
  cons <- rbind(cons,
                result$solution)
  dir <- c(dir, "<=")
  rhs <- c(rhs, 5)
  result <- lp("max", obj, cons, dir, rhs, all.bin = TRUE)
  team <- df_new[which(result$solution == 1), ]
  results[i+1, ] <- c(team$name, sum(team$avg_points_per_game), sum(team$salary))
}
```

```{r}
df <- dk_nyr_car |> 
  filter(position != "G" | name %in% c("Antti Raanta", "Igor Shesterkin")) |>  
  mutate(new_pts = ifelse(roster_position == "CPT",
                          1.5*avg_points_per_game,
                          avg_points_per_game))
obj <- df$new_pts

players_con <- t(model.matrix(~ name + 0, df))
cons <- rbind(rep(1, nrow(df)), #6 players
              df$salary, #<= 50000
              t(model.matrix(~ team_abbrev + 0, df)), # >= one player / team
              ifelse(df$roster_position == "CPT", 1, 0), # one captain
              players_con)

dir <- c("=", "<=", ">=", ">=", "=", rep("<=", dim(players_con)[1]))
rhs <- c(6, 50000, 1, 1, 1, rep(1, dim(players_con)[1]))
result <- lp("max", obj, cons, dir, rhs, all.bin = TRUE)
team <- df[which(result$solution == 1), ] |> 
    select(name, position, salary, team_abbrev, roster_position, 
           avg_points_per_game)
team
```

## Lab: Scraping Odds Data to Build Models
## Conclusion

## Exercises

