---
title: "Chapter Six -- Basketball"
author: "Ryan Elmore"
date: "`r Sys.Date()`"
output:
  html_document:
    df_print: paged
  pdf_document: default
---

```{r knitr-options, include = FALSE}
knitr::opts_chunk$set(fig.align="center",
                      warning = FALSE,
                      message = FALSE,
                      comment = NA)
```

```{r}
library(tidyverse)
library(ggplot2)
library(lubridate)
library(gganimate)
library(nbastatR)
library(jpeg)
library(grid)
library(hexbin)
library(gifski)
library(png)
library(mgcv)
```

```{r data}
games <- readRDS("../data/nba-games-2021.rds")
ff <- readRDS("../data/four-factors-team-2021.rds")
adv <- readRDS("../data/advanced-team-2021.rds")
```

## Problem 1

Mutate the `games` (`nba-games-2019.rds`) table to include a column for Free
Throw Rate (FTR). You will need to use the `fgaTeam` and `ftaTeam` columns in 
the table. Plot your new variable against `rateFTA` from the `ff` table 
(`four-factors-team.rds`).

## Answer 1

```{r}
df_ff_games <- merge(games, ff) %>% 
  dplyr::mutate(., ftr = ftaTeam/fgaTeam)
p <- ggplot(data = df_ff_games,
            aes(x = ftr, y = rateFTA))
p + geom_line() +
  theme_bw()
```

## Problem 2

```{r}
game_summary <- games %>% 
  dplyr::select(., dateGame, idGame, slugTeam, slugMatchup, locationGame,
                outcomeGame, plusminusTeam) 
```

Compute the average pace for the Denver Nuggets when they play at home and when
they play away from Denver. The data are in the `adv` table and you will need
to join this with the `game_summary` object defined above. 

## Answer 3

```{r}
tmp <- inner_join(game_summary, adv) %>%
  group_by(slugTeam, locationGame) %>% 
  summarize(avg_pace = mean(pace)) %>% 
  arrange(desc(avg_pace))
```

## Problem 4

Construct a new data frame (call it `df_ff`) that has game id, game date, team,
the four factors, offensive rating (ortg), defensive rating (drtg), win/loss,
home/away, and plusminusTeam. Create an is_win variable to be TRUE if it was a
win and FALSE if an L. 

## Answer 4

```{r}
df_ff <- inner_join(adv, ff) %>%
  dplyr::inner_join(game_summary) %>% 
  dplyr::select(dateGame, idGame, slugTeam, pctOREB, pctTOVTeam, pctEFG,
                rateFTA, ortg, drtg, outcomeGame, locationGame,
                plusminusTeam) %>%
  dplyr::mutate(is_win = ifelse(outcomeGame == "W", T, F))
  
```

## Problem 5

Compute the correlation matrix between the four factors. Is the result
surprising? Why or why not?

## Answer 5

```{r}
cor(select(df_ff, pctEFG, pctOREB, pctTOVTeam, rateFTA))
```

## Problem 6

Filter the `adv` data to get both teams information from the April 24, 2021 game
between the Los Angeles Lakers and the Dallas Mavericks. Compute the z scores
for offensive and defensive ratings for both teams relative to the league
distributions.

## Answer 6
```{r}
lal_dal_adv <- merge(game_summary, adv) %>% 
  dplyr::filter(dateGame == ymd("2021-04-24"),
                slugTeam %in% c("LAL", "DAL"))
lal_dal_adv$slugTeam
pnorm((lal_dal_adv$ortg - mean(adv$ortg))/sd(adv$ortg))
pnorm((lal_dal_adv$drtg - mean(adv$drtg))/sd(adv$drtg))
```

## Problem 7

Using the `df_ff` data set, run a multiple regression model using the four 
factors and offensive rating to predict plusminusTeam. What are the predicted 
values of plusminusTeam for both teams from Problem 6? 

## Answer 7

```{r}
ff_lm <- lm(plusminusTeam ~ pctOREB + pctTOVTeam + pctEFG + rateFTA + ortg,
            data = df_ff)
```

## Answer 7

```{r}
ff_lm <- lm(plusminusTeam ~ pctOREB + pctTOVTeam + pctEFG + rateFTA + ortg,
            data = df_ff)
summary(ff_lm)
```


### Problem 1

```{r, eval = T}
df <- readRDS("../data/nuggets-2015-21.rds")
```

Use the data related to the Nuggets shots from 2015 through 2020 given here and
create a barchart showing the total number of observations by season. Show the 
zone (`zoneBasic`) as a color within each bar. 

### Answer 1

```{r}
p <- ggplot(data = df,
            aes(x = slugSeason, fill = zoneBasic))
p + geom_bar() +
  scale_fill_brewer("", palette = "Dark2")
```

### Problem 2

Who took the most shots in each season?

### Answer 2

```{r}
group_by(df, slugSeason, namePlayer) %>%
  summarize(n = n()) %>%
  filter(n == max(n))
```

### Problem 3
Compute the shot chart for Denver's 2020-21 season. Use points colored by the
zone. 

### Answer 3
```{r}
court <- rasterGrob(readJPEG("../fig/nba_court.jpg"), 
                    width = unit(1, "npc"),
                    height = unit(1, "npc"))

p <- ggplot(data = df %>% filter(slugSeason == "2020-21"),
            aes(x = locationX, y = locationY, col = zoneBasic))
p + annotation_custom(court, -250, 250, -50, 420) +
  geom_point(alpha = .15) +
  scale_color_brewer("Zone", palette = "Set1") +
  coord_fixed() +
  xlim(250, -250) +
  ylim(-50, 420)
```

### Problem 4
Add a layer to the previous figure showing points shaped by typeEvent (make
or miss the shot). 

```{r}
p <- ggplot(data = df %>% filter(slugSeason == "2020-21"),
            aes(x = locationX, y = locationY, col = zoneBasic,
                shape = typeEvent))
p + annotation_custom(court, -250, 250, -50, 420) +
  geom_point(alpha = .15) +
  scale_color_brewer("Zone", palette = "Set1") +
  coord_fixed() +
  xlim(250, -250) +
  ylim(-50, 420)
```

### Problem 5

Compare Michael Porter Jr.'s shots from 2020 and 2021. Create a 2x2 facet plot
showing his makes/misses by zone for both years. 

```{r}
p <- ggplot(data = df %>% filter(yearSeason %in% c(2020, 2021),
                                 namePlayer == "Michael Porter Jr."),
            aes(x = locationX, y = locationY, col = zoneBasic,
                shape = typeEvent))
p + annotation_custom(court, -250, 250, -50, 420) +
  geom_point(alpha = .15) +
  facet_grid(yearSeason ~ typeEvent) +
  scale_color_brewer("Zone", palette = "Set1") +
  coord_fixed() +
  xlim(250, -250) +
  ylim(-50, 420)

```


### Problem 1

Create an animation showing how the team's shots evolved over the six years in
this data set. Remove the shape aesthetic. What do you observe to have happened 
over the years?

### Answer 1

```{r}
ggplot(data = df,
            aes(x = locationX, y = locationY, col = zoneBasic)) +
  annotation_custom(court, -250, 250, -50, 420) +
  geom_point(alpha = .15) +
  scale_color_brewer("Zone", palette = "Set1") +
  coord_fixed() +
  xlim(250, -250) +
  ylim(-50, 420) +
  transition_manual(slugSeason) +
  labs(title = 'Season: {current_frame}')
```

### Problem 2

Construct an animated bar chart showing the count of each type of shot, whether
it was made (`typeEvent`), over each season. The bars should represent each
zone (`zoneBasic` variable) for each season. 

### Answer 2

```{r}
ggplot(data = df %>% filter(zoneBasic != "Backcourt"),
            aes(x = zoneBasic, fill = typeEvent)) +
  geom_bar() +
  coord_flip() +
  scale_fill_brewer("", palette = "Set1") +
  transition_states(slugSeason) +
  labs(title = 'Season: {closest_state}')
```

### Problem 3

Download the shot charts for the 2021 Golden State Warriors and construct a 
heat map showing the probability of Stephen Curry making shots. 

### Answer 3
```{r}
df <- teams_shots(teams = "Golden State Warriors", seasons = 2021)
player <- "Stephen Curry"
steph_df <- filter(df, namePlayer == player)
make_gam <- gam(isShotMade ~ s(locationX, locationY), #Model for P(make; x & y loca)
                family = binomial, #logistic regression GAM
                data = steph_df)

# Find predicted probabilities over a 50 x 50 grid
make_predict_df <- expand.grid(locationX = seq(-250, 250, length = 50), 
                               locationY = seq(-50, 420, length = 50))

# Get the predicted values from the model and convert to probability values:
make_preds <- predict(make_gam, 
                      newdata = make_predict_df,
                      type = "response")
make_predict_df <- make_predict_df %>%
  mutate(., make_prob = make_preds)

p <- ggplot(data = make_predict_df,
            aes(x = -locationX, y = locationY, fill = make_prob))
p + annotation_custom(court, -250, 250, -50, 420) +
  geom_tile(alpha = .75) +
  scale_fill_distiller("P(Make)", palette = "Spectral") +
  coord_fixed() +
  scale_x_continuous("", breaks = NULL, limits = c(-250, 250)) +
  scale_y_continuous("", breaks = NULL, limits = c(-50, 420))

```


```{r}
df <- readRDS("../data/nba_stats.rds")
```

### Problem 1

Find the game in which the highest offensive rating in this data set was
achieved. What team (`slugTeam`) had this rating?

### Answer 1

```{r}
dplyr::arrange(df, desc(ortg))
```

### Problem 2

What team has the lowest average offensive rating in this data set? 

### Answer 2

```{r}
df %>% dplyr::group_by(., slugTeam) %>% 
  dplyr::summarize(., avg = mean(ortg)) %>% 
  dplyr::arrange(desc(-avg))
```

### Problem 3

What team (`slugTeam`) has the largest absolute difference between their home 
and away average offensive rating? 

### Answer 3

```{r}
df %>% 
  dplyr::group_by(., slugTeam, locationGame) %>% 
  dplyr::summarise(., avg = mean(ortg)) %>% 
  dplyr::ungroup() %>% 
  tidyr::pivot_wider(., names_from = locationGame, values_from = avg) %>% 
  dplyr::mutate(., diff = abs(H - A)) %>% 
  dplyr::arrange(., desc(diff))
```

### Problem 4

Compute a logistic regression of offensive rating ($X_1$ variable) and the game
location ($X_2$) as predictor variables of the probability of winning the game. 
You will need to use `as.factor(outcomeGame)` or something similar as your Y 
variable. Predict the probability of winning the game for a team with an 
offensive rating equal to 115 (`ortg = 115`) playing away (`locationGame = "A"`).
Round your answer to the nearest hundreth. 

```{r}
four_glm <- glm(as.factor(outcomeGame) ~ ortg + locationGame,
                family = "binomial",
                data = df)
predict(four_glm, 
        newdata = data.frame(outcomeGame = NA, 
                             ortg = 115, 
                             locationGame = "A"),
        type = "response")
```

```{r}
summary(four_glm)
```

