---
title: "Chapter XXX -- Soccer"
author: "Ryan Elmore"
date: "`r Sys.Date()`"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r knitr-options, include = FALSE}
knitr::opts_chunk$set(fig.align="center",
                      warning = FALSE,
                      message = FALSE,
                      comment = NA)
```

```{r, eval=F}
remotes::install_github("rtelmore/ISAR")
```

```{r libs, include=FALSE}
library(dplyr)
library(ggplot2)
library(lubridate)
library(rvest)
library(janitor)
library(scales)
library(readr)
library(nhlapi)
library(purrr)
library(sportyR)
library(ggrepel)
library(ggdendro)
library(ISAR)
theme_set(theme_bw())
```

# Soccer Data

In this chapter, we will explore several unsupervised learning methods, e.g. 
hierarchical clustering, using several hockey data sets. 

## FB Reference 

In October 2022, the Sports Reference website announced a partnership with 
Stats Perform Opta to dramatically increase the quantity and quality of soccer, 
or football, data that are available on its site. Soccer data are now hosted at
https://fbref.com/en/ and includes data on most major professional leagues 
throughout the world at a team and player level. More details related to this 
new partnership can be found here. 

https://www.sports-reference.com/blog/2022/10/fbref-leagues-🇵🇹-leagues-🇧🇷-leagues-🇲🇽-expanded-womens-and-mens-data-new-data-partner/

In this chapter, we will focus on using soccer data to cluster both teams and 
individual players using data from fbref.com. 

## Expected Goals

Similar to hockey, soccer is notoriously low-scoring and there is 
certainly an element of luck in scoring a goal. Therefore, one of the main 
developments in the soccer-analytics space is the introduction of Expected Goals, 
or xG for short. There are various models to compute xG, but at a high level xG is
simply the probability of scoring a goal from a given spot on the field when a 
shot is taken.

## Player Statistics

We will initially focus our attention on clustering individual players from the 
American NWSL. The NWSL is National Women's Soccer League and is the largest
women's soccer league in North America. In the 2022 season there were twelve teams
and the Portland Thorns were crowned champion. 

We downloaded a csv file of the NWSL player statistics
from the following fbref.com website and the data are also given in the book's R 
package as `nwsl_player_stats`. Note that the column, or variable, names are defined 
on the Football Reference website. https://fbref.com/en/comps/182/stats/NWSL-Stats


Here's how the data were obtained. Note, they are now located in the `ISAR` package
under the `nwsl_player_stats` object. 

```{r}
nwsl_player_stats <- read.csv("../data/nwsl-players.csv", header = T) |> 
  janitor::clean_names() |> 
  dplyr::select(c(2:6, 8:10, 28:32)) |> 
  dplyr::rename(xGp90 = x_g_1,
                xAp90 = x_ag_1,
                xGxAp90 = x_g_x_ag,
                npxGp90 = npx_g_1,
                npxGxAp90 = npx_g_x_ag_1)
saveRDS(nwsl_player_stats, "../data/nwsl_player_stats.rds")
```

### Exploratory Data Analysis

We will begin by exploring expected goals. In particular, let's look at xG by 
age in the NWSL. We will use the `x_g_1` column as our xG value because it is the
column representing the normalized xG value, normalizing xG per 90 minutes played. 

```{r}
p <- ggplot(data = nwsl_player_stats |> 
              dplyr::filter(pos != "GK"),
            aes(x = age, y = xGp90))
p + geom_point(alpha = .5) +
  labs(x = "age", y = "expected goals (xG)") +
  theme_bw()
```
One thing that we might want to do in plotting sports data is to identify the 
players that we are plotting. In this case, we can use the `ggrepel` package to 
quickly add labels to certain points using the `label` aesthetic and the
`geom_text_repel()` function. Suppose we want to identify players who are less 
than 20 years of age with a positive xG value, or have an xG value greater than 
0.6. 

```{r}
p <- ggplot(data = nwsl_player_stats |> 
              dplyr::filter(pos != "GK") |> 
              dplyr::mutate(new_player = ifelse((age < 20 & xGp90 > 0) | xGp90 > 0.6,
                            player, "")),
            aes(x = age, y = xGp90, label = new_player))
p + geom_point(alpha = .5) +
  labs(x = "Age", y = "Expected goals (xG)") +
  geom_text_repel() +
  theme_bw()
```

## Clustering

In this chapter we will focus on several analytical tools that fall under the  
statistical, or machine learning umbrella of unsupervised learning. 

### K-Means

#### With Age

First let's look at a clustering based on age, min, and xGxAp90. We are going to
consider three clusters initially. 

```{r}
nwsl_cluster_df <- nwsl_player_stats |> 
  dplyr::mutate(player_team = paste(player, squad, sep = "_")) |> 
  dplyr::select(player, squad, player_team, age, min, starts, xGp90, xAp90, xGxAp90, 
                npxGp90, npxGxAp90) |> 
  dplyr::filter(xGxAp90 < 1.5) |> 
  na.omit()

set.seed(9019)
km <- kmeans(dplyr::select(nwsl_cluster_df, age, min, xGxAp90) |> 
               na.omit(),
             centers = 3)
```

How can we visualize these clusters? Let's plot age versus xGxAp90 and color these
points by their clusters. 

```{r}
nwsl_cluster_df <- dplyr::mutate(nwsl_cluster_df, k_means = km$cluster)
p <- ggplot(data = nwsl_cluster_df,
            aes(x = age, y = xGxAp90, col = as.factor(k_means)))
p + geom_point() +
  scale_color_brewer("cluster", palette = "Dark2")
```

```{r}
p <- ggplot(data = nwsl_cluster_df,
            aes(x = min, y = age, col = as.factor(k_means)))
p + geom_point() +
  scale_color_brewer("cluster", palette = "Dark2")
  

```

#### Without Age

```{r}
nwsl_cluster_df <- nwsl_player_stats |> 
  dplyr::mutate(player_team = paste(player, squad, sep = "_")) |> 
  dplyr::select(player, squad, player_team, age, min, starts, xGp90, xAp90, xGxAp90, 
                npxGp90, npxGxAp90) |> 
  dplyr::filter(xGxAp90 < 1.5) |> 
  na.omit()

set.seed(9019)
km <- kmeans(dplyr::select(nwsl_cluster_df, min, starts, xGxAp90) |> 
               na.omit(),
             centers = 4)

nwsl_cluster_df <- dplyr::mutate(nwsl_cluster_df, clusters = km$cluster)

```

```{r}
p <- ggplot(data = nwsl_cluster_df,
            aes(x = min, y = starts, col = as.factor(clusters)))
p + geom_point() +
  scale_color_brewer("cluster", palette = "Dark2")
  

```

```{r}
p <- ggplot(data = nwsl_cluster_df,
            aes(x = starts, y = xGxAp90, col = as.factor(clusters)))
p + geom_point() +
  scale_color_brewer("cluster", palette = "Dark2")
  

```

### Hierarchical Clustering

```{r}
nwsl_cluster_df <- nwsl_player_stats |> 
  dplyr::mutate(player_team = paste(player, squad, sep = "_")) |> 
  dplyr::select(player, squad, player_team, age, starts, xGp90, xAp90, xGxAp90, 
                npxGp90, npxGxAp90) |> 
  na.omit()
row.names(nwsl_cluster_df) <- nwsl_cluster_df$player_team
hc <- hclust(dist(nwsl_cluster_df), "average")
p <- ggdendrogram(hc, rotate = FALSE, size = 2)
p
```




### Gaussian Mixtures

## Exercises

## Case Study

## Lab